Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-20 paper/paper.tex
Check in EOD

\newif\ifdraft
\draftfalse
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
\newcommand{\nathan}[1]{\ifdraft{\color{green}[{\bf Nathan says}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

@@

\begin{figure}
\begin{lstlisting}
f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
f (a , b) = inl (a (*@$\otimes$@*) b)
\end{lstlisting}\begin{mathpar}
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{\stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once. That is, we want to prevent the following
erroneous derivations,
\[
  a : \literal a , b : \literal b \not \vdash a : \literal a \qquad a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a
\]

@@

\begin{figure}
\begin{mathpar}
\footnotesize
\inferrule
{
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {
          \inferrule
          {~}
          {\cdot \vdash \cons : \ltonl {({\literal a} \lto {\literal a}^* \lto {\literal a}^*)}}
        }
        {\cdot \vdash \cons : {\literal a} \lto {\literal a}^* \lto {\literal a}^*}
        \\
        \inferrule
        {~}
        {a : {\literal a} \vdash a : {\literal a}}
      }
      {a : a \vdash \cons~a : {\literal a}^* \lto {\literal a}^*}
      \\
      \inferrule
      {
        \inferrule
        {~}
        {\cdot \vdash \nil : \ltonl {({\literal a}^*)}}
      }
      {\cdot \vdash \nil : {\literal a}^*}
    }
    {a : {\literal a} \vdash \cons~a~\nil : {\literal a}^{*}}
    \\
    \inferrule
    {~}
    {b : {\literal b} \vdash b : {\literal b}}
  }
  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil , b) : {\literal a}^{*} \otimes {\literal b}}
}
{a : \literal a , b : \literal b \vdash \texttt{g} := \inl (\cons~a~\nil , b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
\end{mathpar}
\begin{lstlisting}
g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
g (a , b) = inl (cons a nil (*@$\otimes$@*) b)
\end{lstlisting}
\caption{\stringquote{ab} is parsed by
  $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
\label{fig:kleenestarderivation}
\end{figure}

@@

\begin{figure}
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {\cdot \vdash \nil : \A^*}
    \\
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {~}
        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons~a_1 (\cons~a_2~a_{*}) : \A^*}
      }
      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2}
        {a'} {\cons~a_1 (\cons~a_2~a_{*}) : \A^*}}
    }
    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
  }
  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
\end{mathpar}
\begin{lstlisting}
h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
h nil = nil
h (cons (a1 (*@$\otimes$@*) a2) as) = cons a1 (cons a2 (h as))
\end{lstlisting}
\caption{A parse transformer for abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

\paragraph{Distributivity}
First, we need that additive conjunction \emph{distributes}
over additive disjunction --- e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
general form, the axiom says that the definable function
${\bigoplus\limits_{f : \prod_{x:X}
    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
inverse.

\steven{Maybe this detail should go closer to the actual example
We use distributivity to encode the lookahead of the LL(1) parser in
\cref{thm:exp-parser}. At parse time, an LL(1) automaton peeks at the next
character of a string and uses the peeked character to inform its next
transition. Our automaton from \cref{fig:binop-inductive} chooses which state to
transition to dependent on if the next character is $\stringquote{)}$.

\paragraph{Disjointness of Constructors}
Second, we need that the different constructors of
$\bigoplus$ are disjoint. We add this by adding the axiom that for any
$A : X \to L$ and $x \neq x' : X$ there is a function
$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
is empty.

\pedro{imo, since we are providing an Agda implementation, we can
  be looser with the explanation of the meaning of the disjointness axioms.
  As it stands, it's a bit hard do follow along. Is there a simpler example
  we could point to?}

@@

%% \paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$.
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parses of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
  each $A(x)$ is unambiguous.
\end{lemma}
From the first principle, we can prove that $\String$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
defined as a Kleene star, we can easily show that $\String \cong \I
\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
the second principle we have that $\I$ and $\Char$ and each literal
$\literal c$ are unambiguous as well.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\String \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string doesn't
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.
\begin{definition}
  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
    A \oplus A_{\neg})}$, where $A_{\neg}$ is a linear type that is
  \emph{disjoint} from $A$ in that we can implement a function
  $\uparrow{(A \& A_{\neg} \lto 0)}$.
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to parse the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse. Fortunately, we will see that
deterministic automata naturally support such a notion with no
additional effort: the negative grammar is simply the grammar for
traces that end in a rejecting state. This follows from the following
principle, a consequence of our axiom that $\oplus$ constructors are
disjoint.
\begin{lemma}[\agdalogo]
  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
  disjoint.
\end{lemma}

@@

>>>>

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
\newcommand{\nathan}[1]{\ifdraft{\color{green}[{\bf Nathan says}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

\newtheorem{axiom}{Axiom}[section]
\newtheorem{construction}{Construction}

@@

\begin{figure}
\begin{lstlisting}
f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
f (a , b) = inl (a , b)
\end{lstlisting}\begin{mathpar}
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{\stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once. That is, we want to prevent the following
erroneous derivations,
\[
  a : \literal a , b : \literal b \not \vdash a : \literal a \qquad a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a
\]

@@

\begin{figure}
\begin{mathpar}
\footnotesize
\inferrule
{
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {
          \inferrule
          {~}
          {\cdot \vdash \cons : \ltonl {({\literal a} \lto {\literal a}^* \lto {\literal a}^*)}}
        }
        {\cdot \vdash \cons : {\literal a} \lto {\literal a}^* \lto {\literal a}^*}
        \\
        \inferrule
        {~}
        {a : {\literal a} \vdash a : {\literal a}}
      }
      {a : a \vdash \cons~a : {\literal a}^* \lto {\literal a}^*}
      \\
      \inferrule
      {
        \inferrule
        {~}
        {\cdot \vdash \nil : \ltonl {({\literal a}^*)}}
      }
      {\cdot \vdash \nil : {\literal a}^*}
    }
    {a : {\literal a} \vdash \cons~a~\nil : {\literal a}^{*}}
    \\
    \inferrule
    {~}
    {b : {\literal b} \vdash b : {\literal b}}
  }
  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil , b) : {\literal a}^{*} \otimes {\literal b}}
}
{a : \literal a , b : \literal b \vdash \texttt{g} := \inl (\cons~a~\nil , b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
\end{mathpar}
\begin{lstlisting}
g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
g (a , b) = inl (cons a nil , b)
\end{lstlisting}
\caption{\stringquote{ab} is parsed by
  $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
\label{fig:kleenestarderivation}
\end{figure}

@@

\begin{figure}
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {\cdot \vdash \nil : \A^*}
    \\
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {~}
        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons~a_1 (\cons~a_2~a_{*}) : \A^*}
      }
      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2}
        {a'} {\cons~a_1 (\cons~a_2~a_{*}) : \A^*}}
    }
    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
  }
  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
\end{mathpar}
\begin{lstlisting}
h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
h nil = nil
h (cons (a1 , a2) as) = cons a1 (cons a2 (h as))
\end{lstlisting}
\caption{A parse transformer for abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

\paragraph{Distributivity}
First, we need that additive conjunction \emph{distributes}
over additive disjunction --- e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most general form,

\begin{axiom}[Distributivity]
\label{ax:dist}
For any $A : (x : X) \to Y x \to L$, the function

\[
 \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \ltonl {\left(  \LinSigTyLimit{f}{\prod\limits_{x : X} . Y (x)}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
\LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}} \right)}
\]

has an inverse.
\end{axiom}

Our primary use of distributivity is to define an equivalence that
expresses a linear type $A$ as a sum over which character it starts with, if any,
\(
A \cong \left(A \& I \right) \oplus \LinSigTyLimit{c}{\Sigma_0}{A \& \left( \literal{c} \otimes \top \right)}
\). We use this equivalence when building a parser for the traces of the lookahead
automaton in \cref{fig:binop-inductive}.

\begin{corollary}
  \label{cor:binary-mono}
  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms.
\end{corollary}

\paragraph{Disjointness of Constructors}
Second, we need that the different constructors of
$\bigoplus$ are disjoint. That is, we want to enforce that $\sigma~x~a$ and
$\sigma~x'~a'$ of type $\LinSigTyLimit{x}{X}{A\,x}$ are not equal whenever $x \neq x'$.
However, because these are linear terms we cannot state their disequality
directly. Instead, we encode the disequality via a function out of an equalizer,

\begin{axiom}[$\sigma$-Disjointness]
\label{ax:disjointness}
For any $A : X \to L$ and $x \neq x' : X$ there is a function,
\[
  \uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
  \lto 0)
\]
  where $b: A(x) \& A(x')$. That is, the grammar of pairs of an
$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
is empty.
\end{axiom}

\steven{need to get cref to cite this as a lemma}
We use $\sigma$-disjointness in \cref{lem:unambig-to-disjoint} to proof that
unambiguous sums have disjoint summands.

% \pedro{imo, since we are providing an Agda implementation, we can
%   be looser with the explanation of the meaning of the disjointness axioms.
%   As it stands, it's a bit hard do follow along. Is there a simpler example
%   we could point to?}

@@

%% \paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$.
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parses of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
  As a consequence of \steven{fix cref} \cref{cor:binary-mono}, if a binary disjunction $A \oplus B$ is unambiguous then
  $A$ and $B$ are each unambiguous.
\end{lemma}
From the first principle, we can prove that $\String$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
defined as a Kleene star, we can easily show that $\String \cong \I
\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
the second principle, we have that $\I$ and
$\Char$ and each literal
$\literal c$ are unambiguous as well.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\String \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string doesn't
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.

\begin{definition}
  Linear types $A$ and $B$ are \emph{disjoint} if there is a function
  $\ltonl {\left( A \& B \lto 0 \right)}$.
\end{definition}

\begin{definition}
  A parser for a linear type $A$ is the choice a type $A_{\neg}$ disjoint from
  $A$ and function $\uparrow{(\String \lto
    A \oplus A_{\neg})}$.
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to parse the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse.

Fortunately, we will see that
deterministic automata naturally support such a notion with no
additional effort: the negative grammar is simply the grammar for
traces that end in a rejecting state. This follows from the following
principle, a consequence of \cref{ax:disjointness}.

\steven{
\begin{lemma}[\agdalogo]
  \label{lem:unambig-to-disjoint}
  If $\LinSigTyLimit{x}{X}{A\,x}$ is unambiguous, then for $x \neq x'$, $A\,x$ and $A,\y$ are
  disjoint.

  In particular, if the binary product $A \oplus A_{\neg}$ is unambiguous, then
  $A$ and $A_{\neg}$ are disjoint.
\end{lemma}
}

@@

\steven{Maybe this detail should go closer to the actual example
We use distributivity to encode the of the LL(1) parser in
\cref{thm:exp-parser}. At parse time, an LL(1) automaton peeks at the next
character of a string and uses the peeked character to inform its next
transition. Our automaton from \cref{fig:binop-inductive} chooses which state to
transition to dependent on if the next character is $\stringquote{)}$.
